---
title: "1- Introduction to the R language"
author: "Alex Sanchez, Miriam Mota, Ricardo Gonzalo and 
        Mireia Ferrer"
date: "Statistics and Bioinformatics Unit. Vall d'Hebron Institut de Recerca"
output:
  beamer_presentation:
    theme: "Copenhagen"
    colortheme: "dolphin"
    fonttheme: "structurebold"
  slide_level: 2
footer: "R for Data Science"

---

```{r setLicense, child = 'license.Rmd'}
```

# Introduction to R

## Outline

- A first contact with R & Rstudio. 
    + How does one work with R
- A primer of data import
    + Reading data into R
- A primer of communication
    + R Notebooks and RMarkdown


# A first contact with R, Rstudio and the `tidyverse`

## [What is R?](https://www.r-project.org/about.html)

- R is a _language and environment_ for statistical computing and graphics.

- R provides a wide variety of statistical and graphical techniques, and is highly extensible.


- It compiles and runs on a wide variety of UNIX platforms and similar systems Windows and MacOS.

## R PRO's (why you are here!)

- The system is 
    + free (as in _free beer_)
    + It's platform independent
    + It is constantly improving (2 new versions/year)
    
- It is a statistical tool
    + Implements almost every statistical method that exists
    + Great graphics (Examples)
    + Simple reporting tools
    + Also state-of-the-art in Bioinformatics through the [Bioconductor Project](http://bioconductor.org).

- Programming language
    + Easy to automate repetitive tasks (Example_1.1)
    + Possibility to create user friendly web interfaces  with a moderate effort. (Examples)
    
## R CON's

- R is mainly used issuing commands from a console
    + less user friendly than almost any other statistical tool you may know.
    
- Constantly having new versions may affect our projects

- Not necessarily the best language nor suitable for every existing task

## How is R used

- Traditionally R was used from an Operating System console ("Terminal")

- This is an intimidating approach for many users

- A variety of options exist to decrease the learning curve.

    + Use a supportive development environment such as **Rstudio**
    + Use an interface to Statistical tools, such as **Rcommander** or ::DeduceR** allowing to concetrate an Statistics, not in commands.
    
## A raw R console in linux

\begin{figure}
\includegraphics[width=0.85\linewidth]{images/RConsole.png}
\end{figure}

## An "enhanced" console: Rstudio

\begin{figure}
\includegraphics[width=0.85\linewidth]{images/RStudio.png}
\end{figure}

## Something that is not a console: Rcommander

\begin{figure}
\includegraphics[width=0.85\linewidth]{images/RCommander.png}
\end{figure}

# Using R

## Commands, Objects and Functions

- Shortly, using R consists of 
    + Working with *objects* using *commands* and *functions*

\begin{figure}
\includegraphics[width=0.85\linewidth]{images/basicConcepts.png}
\end{figure}

## Variables and data types

- Data managed in R ...
    + is stored as *variables*
- Variables can be of distinct types
    + Numerical
        - numeric (13.7)
        - int (3)
    + Character
        - "R is cute"
    + Factors
        - A,B,C,D
        - WT, Mut
        

## R packages

- R can be used for many different types of data processing and analysis from distinct fields, besides statistics such as Ecology, Omics Sciences, Psychology etc.
- All these capabilities are not present from the begining because most of them will never be used by most users.
- Instead, thay can be added when needed by 
    (i) installing and 
    (ii) loading the appropriate packages.

## Installing and loading packages
 
We want to analyze some data using cox proportional hazards model.
```{r eval=FALSE}
res.cox <- coxph(Surv(time, status) ~ sex, data = lung)
```
`Error in coxph(Surv(time, status) ~ sex, data = lung) : could not find function "coxph"`

We need to install and load the package before we can use it.
```{r eval=FALSE}
install.packages("survival")
library(survival)
res.cox <- coxph(Surv(time, status) ~ sex, data = lung)
```

## The `tidyverse`

- The tidyverse is an opinionated collection of R packages designed for data science. 
- All packages share an underlying design philosophy, grammar, and data structures. 
- The complete tidyverse collection can be installed with:
```{r eval=FALSE}
install.packages("tidyverse")
```
- [https://www.tidyverse.org/](https://www.tidyverse.org/)


# Getting data into R

## Importing data with Rstudio

- The easiest way to get data into R is to click on the `Ìmport Datasets` button.
- Alternatively R code can be written using functions from `Base R` or the `tidyverse`
    + `Base R` functions start with `read.`: `read.table`, `read.csv`
    + `tidyverse` functions start with `read_`: `read_delim`, `read_csv` or `read_excel`

## Reading Excel or csv files

- Files can be read from any location, let it be a physical support or a web site.
- To read files from disk be sure to indicate their location.
- Alternatively the default working directory can be set to the folder where the file is located.

- Assume files `Diabetes.xls` and `Osteoporosis.csv` have been downloaded from url \url{https://github.com/uebvhir/uebvhir.github.io/blob/master/datasets} to a sub-folder named `datasets` 
- Start setting the default directory to the folder where you have saved the `datasets` folder.
    + `Session --> Set Working directory --> To source file location...`

- Import the `diabetes.xls` and the `osteoporosis.csv` file with the default options

## Reading Excel or csv files (continued)

The code generated for reading the files can be reused any time changing the file name if needed.

```{r}
# Read Excel file
library(readxl)
diabetes <- read_excel("datasets/diabetes.xls")
```


## Reading text files

- Text files may require that more information is provided about delimiters, decimal dign, locale (language) or page encoding (UTFB for Mac or Linux vs ISO-8859-1 for Windows).
- All options can be selected from the rstudio importer

```{r message=FALSE}
# Read csv file
library(readr)
osteoporosis <- read_delim("datasets/osteoporosis.csv", 
     "\t", escape_double = FALSE, locale = locale(date_names = "es", 
     decimal_mark = ",", encoding = "ISO-8859-1"), 
     trim_ws = TRUE)
```

## Interlude: Summarizing data

- Once a dataset is available it is easy to "have a look at it"

```{r eval=FALSE}
head(diabetes)
str(diabetes)
summary (diabetes)
```


# Dynamic output with Rmarkdown

## Reproducible research with R notebooks

- R and Rstudio are strongly involved in promoting [reproducibility](https://en.wikipedia.org/wiki/Reproducibility) and [reproducible research](https://en.wikipedia.org/wiki/Reproducibility#Reproducible_research).
- This is implemented in **R notebooks**
- A notebook combines
    + Natural language text, e.g. describing what we are doing in our own words.
    + R code with the instructions needed to do the data management or the analysis.
    + The output of the analysis

## Creating Notebooks

- A notebook can be created in Rstudio with 
    + `File --> New File --> R Notebook`
- The notebook contains example text and code so it is straightforwoard to adapt it to your analysis.
- To produce an html file with text, code and output:
    - Press the button "Preview"
    - Or Select "Knitr to Html"


# Your turn

## Exercise

- Select a dataset with which you wish to work along the course.
- Read it into R
    + How many variables are there in it
    + What are their types
- Try to summarize it briefly
- Create an R notebook to encapsulate all your steps and share it with somebody.