---
title: "R for Data Science (I): Visualization"
author: "Alex Sanchez, Miriam Mota, Ricardo Gonzalo and 
        Mireia Ferrer"
date: "Statistics and Bioinformatics Unit. Vall d'Hebron Institut de Recerca"
output:
  beamer_presentation:
    theme: "Copenhagen"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    slide_level: 2
footer: "R for Data Science"
editor_options: 
  chunk_output_type: console
---

```{r setLicense, child = 'license.Rmd'}
```

## Outline: Data Exploration

- The Data Science Approach in R
- Data Visualization
- Data Transformation
- Exploratory Data Analysis

## Recall: The Data Science Approach in R 

\begin{figure}
\includegraphics[width=1\linewidth]{images/data-science.png}
\end{figure}

# Data Visualization

## Introduction

> "The simple graph has brought more information to the data analyst’s mind than any other device.” 
>
> `r tufte::quote_footer('--- John Tukey')`

\bigskip

We consider three components of visualization:

1. Aesthetics

2. Facetting

3. Geoms

# Aesthetic mappings

## Datasets in packages

- Packages include example datasets

- Datasets can be loaded by loading the package 

```{r}
library(ggplot2)
data(package="ggplot2")
```

## The `mpg` dataset

- The dataset `mpg` is included in package `ggplot2`

- Obtain some information about it.

```{r eval =FALSE}

?mpg
head(mpg)
str(mpg)
summary(mpg)
dim(mpg)
```

## Scatterplot basics

```{r out.width='70%'}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## Additional information

- Plots can be enhanced by displaying additional information

- **aesthetics** displays it using distinct shapes, colors or sizes.

- **faceting** breaks displays into multiple smaller displays for different subsets.

## Improving plots

For better plot "add" the information to the call

```{r eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy,
                           color = class))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy,
                           alpha = class))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy,
                           shape = class))
```

## Your turn now

- Experiment with colour, size, and shape
aesthetics.

- What’s the difference between discrete or
continuous variables?

- What happens when you combine
multiple aesthetics?

## In summary

|        | Discrete             | Continuous                                |   |
|--------|----------------------|-------------------------------------------|---|
| Colour | Rainbow              | Gradient                                  |   |
| Size   | Disrete size steps   | Linear mapping  radius-value |   |
| Shape  | Different shape each | Doesn't work                              |

# Facets

## Faceting

- Break the visualization in many small plots - 
- Each (sub)plot reflects one of multiple conditions defined by one or more (categorical) variables.
- Useful for exploring conditional relationships or for when there are many data.

## `facet_wrap`: split plots by one variable  

```{r eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) 
+ facet_wrap(~ class, nrow = 2)
```

```{r out.width='70%', echo=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + facet_wrap(~ class, nrow = 2)
```

## `facet_grid`: split plots by two variables

```{r eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) 
+ facet_grid(drv ~ cyl)
```

```{r out.width='70%', echo=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + facet_grid(drv ~ cyl)
```

## Your turn

```{r eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) 
+ facet_grid(. ~ cyl)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) 
+ facet_grid(drv ~ .)
```

# Geometric Objects "Geoms"

## What are "geom"

- A `geom` is the geometrical object that a plot uses to represent data. 

- For example, 

    + Bar charts use bar geoms, 
    + Line charts use line geoms, 
    + Boxplots use boxplot geoms
    + Scatterplots use the point geom !

## Applying `geoms`: How are these plots similar?

```{r echo = FALSE, out.height = "40%", fig.align="default", message = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

## Using `geoms`

- Both plots contain the same x variable and  the same y variable, 
- both describe the same data. 
- Each plot uses a different visual object to represent the data. 
- In ggplot2 syntax, we say that they use different __geoms__.

## Changing `geoms`

- To change the geom in your plot, change the geom function that you add to `ggplot()`

```{r eval = FALSE}
# left
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# right
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
