---
title: "R for Data Science (I): Visualization"
author: "Alex Sanchez, Miriam Mota, Ricardo Gonzalo and 
        Mireia Ferrer"
date: "Statistics and Bioinformatics Unit. Vall d'Hebron Institut de Recerca"
output:
  beamer_presentation:
    theme: "Copenhagen"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    slide_level: 2
footer: "R for Data Science"
editor_options: 
  chunk_output_type: console
---

```{r setLicense, child = 'license.Rmd'}
```

## Outline: Data Exploration

- The Data Science Approach in R
- Data Visualization
- Data Transformation
- Exploratory Data Analysis

## Recall: The Data Science Approach in R 

\begin{figure}
\includegraphics[width=1\linewidth]{images/data-science.png}
\end{figure}

# Data Visualization

## Graphics in the `tidyverse`

- Traditionally graphics in R are relatively complicated because they are based in functions with many parameters.
- Improving a graphic or overimposing distinct plots is also a non-trivial task.

- The `tidyverse approach` provides a distinct way to draw plots which is, at the same time, **intuitive, flexible and powerful**.

- This is made possible because it implements the so-called *grammar of graphics* which was introduced by Hadley Wickam in his paper [A layered grammar of graphics](http://vita.had.co.nz/papers/layered-grammar.pdf).

## The grammar of graphics

- Graphics are treated as a set of elements which can be combined to produce the final plot.

- The idea consists of *working with distinct layers* starting with a first one that sets the data to be plotted.

- Successive layers are added, for instance to change colors, add annotations, overimpose other plots, etc.

## The `ggplot2` package

- This package implements the grammar of graphics within the `tidyverse`.

- The package does not belong to the standard R distribution, so it has to be installed.
    + This can be done when installing the `tidyverse` or separately (only for this package).
    
- Option 1:
```{r eval=FALSE}
install.packages('tidyverse')
```

- Option 2:
```{r eval=FALSE}
install.packages('ggplot2')
```
    
## Creating a plot with `ggplot2`

- A (gg)plot is obtained by combining several elements which  produce distinct layers in the same plot:

>    1. The data to be represented, stored in a data frame. 
>    2. Geometric objects (`geoms`) which define the global aspect of the layer (bars, points, lines...).
>    3. Esthetic attributes (`aesthetics`), visual properties of the geoms such as *position*, *color of line*, *shapes of points*, etc.
>    4. A statistical summary of the data (`stats`) (*counting*, *smoothing*, ...). It is usually associated to the type of geom used.
>    5. `facets` and `scales` allow to visualize different subsets of the data and control the representation in space. 
>    6. Different elements can be included in the graph with the operator `+`.

## Creating a plot in practice

The basic steps to create a plot
are:

1. Create a `ggplot` object providing the data and some `aesthetics`

2. Add one or more `geoms` using the `+` operator to define and shape the plot type.

## Example 1. The data

- First we need the data 
```{r}
library(readxl)
diabetes <- read_excel("datasets/diabetes.xls")
head(diabetes)
```
- We can esily select subsets of our dataset
```{r}
require(dplyr)
diabetesF <- filter(diabetes, tabac=='Fumador')
dim(diabetesF)
```


## Example 1. Build the plot
```{r fig.height=4}
library(ggplot2)
ggplot(diabetes)+
  geom_point(aes(x=edat,y=sbp))
```

## Variations on the theme

- Calls to ggplot can be combined differently

```{r eval=FALSE}
ggplot(data=diabetes,aes(x=edat,y=sbp))+
  geom_point()
```
or 
```{r eval=FALSE}
ggplot()+
  geom_point(data=diabetes,aes(x=edat,y=sbp))
```

## Aesthetics

- In a ggplot aesthetic `aes ()`  refers to *what
we can see*, that is, visual properties of an object.
    + x, y: what goes on the axes
    + color: exterior color
    + ll: color of the interior
    + shape: shape of the points
    + linetype: type of line
    + size: size
    + alpha: transparency (1: opaque; 0: transparent)
    
- Each type of geometry accepts a subset of the possible options.
- One of the most used functions is to define groups through various aesthetics variables or directly with the option on "group".

## Example 2: Aesthetics

```{r fig.height=2}
ggplot(diabetes)+
  geom_bar(aes(x=as.factor(ecg)))
```
```{r fig.height=2}
ggplot(diabetes)+
  geom_histogram(aes(x=bmi),bins=10)

```

## Distinguishing between groups using `aes()` (1)

```{r fig.height=4}
ggplot(diabetes)+
  geom_point(aes(x=edat,y=sbp, col=tabac))
```

## Distinguishing between groups using `aes()` (2)

```{r fig.height=4}
ggplot(diabetes)+
  geom_point(aes(x=edat,y=sbp, col=tabac, shape=chd))
```

## `aes` properties that do not depend on variables

Notice the difference between these plots.

```{r fig.height=2}
ggplot(diabetes)+ 
  geom_point(aes(x=edat, y=sbp), col='red')
```

```{r fig.height=2}
ggplot(diabetes)+
  geom_point(aes(x=edat, y=sbp, col='red'))
```

# Geometric Objects

## Modifying plots by adding `geoms`

- Geometric objects are the actual marks we put on a plot. Examples include:
    + points (`geom_point`, for scatter plots, dot plots, etc)
    + lines (`geom_line`, for time series, trend lines, etc)
    + boxplot (`geom_boxplot`, for, well, boxplots!)

- A plot **must have at least one `geom`**; there is no upper limit.
    - You can add a geom to a plot using the + operator

- You can get a list of available geometric objects using the code below: 
    `help.search("geom_", package = "ggplot2")`
    
## Drawing plots incrementally

- In the console run the follow instructions one after the other

```{r eval=FALSE}
(p <- ggplot(diabetesF, aes(x=edat, y=sbp)))
(p<- p + geom_point())
(p<- p + geom_smooth(method='lm'))
```

## Do not forget boxplots!

```{r, fig.height=4}
(p<- ggplot(diabetes, aes(x=chd, y=tempsviu)) +   
   geom_boxplot())
(p<- p+ 
    ggtitle("Relation between temps viu and chardiac disease"))
```

## Adding labels to your plot

- It is straightforward with the `geom_text()` which accepts a `labels` mapping.
- An alternative is using `geom_label`

```{r fig.height=3}
ggplot(diabetesF, aes(x=edat, y=sbp))+ 
  geom_point() + 
  geom_text(aes(label=chd), size = 3)
```

## ggplot extensions: the `ggrepel` package

Use this package to avoid overlapping of labels and points

```{r eval=FALSE}
install.packages("ggrepel")
```
```{r eval=FALSE}
require(ggrepel)
ggplot(diabetesF, aes(x=edat, y=sbp))+ geom_point() +
  geom_label_repel(aes(label=chd), size = 3)
```

## ggplot extensions: the `ggrepel` package

```{r echo=FALSE, fig.height=4}
require(ggrepel)
ggplot(diabetesF, aes(x=edat, y=sbp))+
  geom_point() + 
  geom_label_repel(aes(label=chd), size = 3)
```


## Exercise I

- The data for this exercise, stored in the file `EconomistData.csv`.
- They consist of Human Development Index and Corruption Perception Index scores for several countries.

1. Create a scatter plot with CPI on the x axis and HDI on the y axis.
2. Color the points blue.
3. Map the color of the the points to Region.
4. Make the points bigger by setting size to 2
6. Map the size of the points to HDI.Rank

# Facets

## Facets

- Faceting is ggplot2 parlance for small multiples

- The idea is to create separate graphs for subsets of data

- ggplot2 offers two functions for creating small multiples:

    + `facet_wrap()`: define subsets as the levels of a single grouping variable
    + `facet_grid()`: define subsets as the crossing of two grouping variables
    
- Faceting facilitates comparison among plots, not just of geoms within a plot

## The `housings` dataset

For te following examples we will use a database on 
housing prices.

```{r}
require(readr)
housing <- read_csv("datasets/landdata-states.csv")
```

## What is the trend in housing prices in each state?

We can start with what we know how to do: map State to color.

```{r eval=FALSE}
p5 <- ggplot(housing, aes(x = Date, y = Home.Value))
p5 + geom_line(aes(color = State)) 
```

## Housing prices trends by states (1)


```{r fig.height=4}
p5 <- ggplot(housing, aes(x = Date, y = Home.Value))
p5 + geom_line(aes(color = State)) 
```

## Housing prices by states (2)

- Visibility of distinct trends depending on state can be improved if we plot each state in a separate graphic. 

```{r eval=FALSE}
p5 <- ggplot(housing, aes(x = Date, y = Home.Value))
(p5 <- p5 + geom_line() +
   facet_wrap(~State, ncol = 10))
```

## Housing prices by states (2)


```{r echo=FALSE, fig.height=6}
p5 <- ggplot(housing, aes(x = Date, y = Home.Value))
(p5 <- p5 + geom_line() +
   facet_wrap(~State, ncol = 10))
```

## Exercise

- Interpret the result of the following instructions:

```{r eval=FALSE}
 ggplot(mtcars,aes(x=wt,y=mpg))+geom_point()+
+ geom_smooth()+
+ facet_grid(as.factor(am)~as.factor(gear))
```

- What happens if we try to separate based on a continuous variable?
- How can this be solved?

# Statistical transforms

## Statistics

- Some plot types (such as scatterplots) do not require transformations-each point is plotted at x and y coordinates equal to the original value. 

- Other plots, such as boxplots, histograms, prediction lines etc. require statistical transformations:

    + for a boxplot the y values must be transformed to the median and 1.5(IQR)
    + for a smoother smother the y values must be transformed into predicted values

## One stat per each geom

- Each geom has a default statistic, but these can be changed.

| geom          	| stat            	|
|---------------	|-----------------	|
| geom_bar()    	| stat_count()    	|
| geom_col()    	| stat_identity() 	|
| geom_pol()    	| stat_identity() 	|
| geom_smooth() 	| stat_smooth()   	|

- The "stat" is an argument of the "geom" and the "geom" is an argument from the "stat".

- Compare the outputs from:
```{r eval=FALSE}
args(geom_histogram)
args (stat_bin)
```

## Seeing the effect of stats (1: default)

```{r, fig.height=4}
p <-ggplot(housing, aes(x = Home.Value))
(p<-p + geom_histogram())
```


## Seeing the effect of stats (2: change values)

```{r, fig.height=4}
p<- ggplot(housing, aes(x = Home.Value))
(p<-p+geom_histogram(stat= 'bin', binwidth=4000))
```

## That's (not) all

- There are other things you can do with your plots.

- Start trying what we have done with your data or check the final exercise in this workshop:

    + [A workshop on R graphics with ggplot2](https://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html)

- And do not forget the cheatsheet!:

    + [ggplot2-cheatsheet-2.1-Spanish.pdf](https://www.rstudio.com/wp-content/uploads/2016/12/ggplot2-cheatsheet-2.1-Spanish.pdf)