---
title: "R for Data Science (V): Programming"
author: "Alex Sanchez, Miriam Mota, Ricardo Gonzalo and 
        Mireia Ferrer"
date: "Statistics and Bioinformatics Unit. Vall d'Hebron Institut de Recerca"
output:
  beamer_presentation:
    theme: "Copenhagen"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    slide_level: 2
footer: "R for Data Science"
editor_options: 
  chunk_output_type: console
---

```{r  include=FALSE}
require(knitr)
opts_chunk$set(
concordance=FALSE, echo=TRUE, cache=FALSE, warning=FALSE, error=FALSE, message=TRUE, fig.height=4, size = "footnotesize")

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

```{r setLicense, child = 'license.Rmd', eval =F}
```

## Outline: Introduction to programming

\vspace{2cm}

- Introduction
- Control structures
- The `apply` family
- User defined functions


\vspace{2.5cm}

\tiny *Based on this presentation: **Programacion en R** by *Antonio Miñarro* Universitat de Barcelona.

## Introduction

- We have introduced R  as a *a language (a tool), to manage and analyze data*.
- It is also a *programming language*
    + It is simple and versatile
    + The user can create new functions that adapt to their
needs
    + It is widely used (2nd most widely used in Data Science)
    + Users provide the community with a high variety of solutions
     + As a programming language it is not, however, very efficient
    
## Example 1: Why may we need programming

- Assume we have the diabetes dataset and want to make a summary of every variable it contains.

```{r, eval=FALSE}
    library(readxl)
    diabetes <- read_excel("datasets/diabetes.xls")
    summary(diabetes)
```
 
 - There are categorical variables but the system cannot recognnize them.
 
---

```{r, echo=FALSE}
    library(readxl)
    diabetes <- read_excel("datasets/diabetes.xls")
    summary(diabetes)
```
---

- A simple solution: Convert text variables into factors.
```{r}
    library(forcats)
    diabetes$mort <- as_factor(diabetes$mort)
    diabetes$tabac <- as_factor(diabetes$mort)
    diabetes$ecg <- as_factor(diabetes$ecg)
    diabetes$chd <- as_factor(diabetes$chd)
```
---
\tiny
```{r}
    summary(diabetes)
```
\normalsize

---

- But how shoulde we proceed if there were dozens or hundreds of variables that need to be changed?
- What if, besides, these variables had different names at every new file?

- The solution consists of providing some way to indicate that "any" character variable is transformed into a factor.
- This will be  an example of a "program",


# Changing the flow of execution

## Scripts are executed "lineally"

- R, as most ordinary programming languages, is executed lineally, that is from the first to last line.
- Sometimes this needs to be changed.
    + Taking alternative flows according to certain conditions
    + Repeating some instructions while certain condition holds, or a fixed number of times,...
    
- This can be acomplished using *Flow control structures*

## Loop controlled by a counter: `for` instruction

- Loops are used in programming to repeat a specific block of code made by one or more instructions.
- Syntax of for loops:

```{r eval=FALSE}
    for (val in sequence)
    {
    statement
    }
```

- Here, `sequence` is a vector and `val` takes on each of its value during the loop. In each iteration, `statement` is evaluated.

## Example of `for` loop

- A `for` loop can be used to change the selected columns in the diabetes dataset.

```{r}
    diabetes <- data.frame(read_excel("datasets/diabetes.xls"))
are_char <- c(2,7,10,11)
    for (i in are_char) {
      diabetes[,i]<-as_factor(diabetes[,i])
    cat(colnames(diabetes)[i], class(diabetes[,i]), "\n")
    }
```
---

```{r}
summary(diabetes)
```

## Exercise

- Create a `for` loop that reads all .csv filenames in your datasets directory (or the directory you decide) and prints the name of the file and the column names in the screen.

## Conditional statements: `if` / `if - else`.

- Conditional statements allow different coding blocks to be executed depending on whether a certain condition is TRUE or FALSE.

- syntax of `if` statement is:

```{r eval=FALSE}
    if (test_expression) {
        statement
      }
```


- If the `test_expression` is TRUE, the `statement` gets executed. But if it's FALSE, nothing happens.
- Here, `test_expression` can be a logical or numeric vector, but only the first element is taken into consideration.
- In the case of numeric vector, zero is taken as FALSE, rest as TRUE.


## Conditional statements: `if - else`.

- syntax of `if-else` statement is:

```{r eval=FALSE}
    if (test_expression) {
        statement_1
    }else{
        statement_2
      }
```

- If the `test_expression` is TRUE, then `statement_1` gets executed. 
- If it's FALSE then `statement_2` gets executed. 

## Exercise

